---
title: "Introducing the PLACES Package"
author: "Brenden Smith"
date: "2023-12-17"
categories: [R, Packages, API, vignette]
description: ""
engine: knitr
execute: 
  warning: false
  eval: true
# image: 
code-fold: show
code-overflow: scroll
toc: true
---

## Introduction
This post contains a brief vignette demonstrating the use of the package `PLACES`.

To begin, we can install `PLACES` from github, then load our packages.

```{r, message = F}
devtools::install_github("brendensm/PLACES")

library(PLACES)
library(dplyr)
library(ggplot2)
```

## Function: `get_measures`
Our first functions allows us to easily view what measures we can query, along with a brief definition. If we run `get_measures`, we must specify a release year. Then we can view the measures in a data frame in the R Studio viewer. If we instead print `measures23`, the pre-loaded data set, we see a preview.

```{r}
head(measures23) %>% tibble()
```

Remember to use the `measureid` when using the next function.

## Function: `get_places`
This function allows us to easily query data that we specify. In the example below, I will get the measure `ACCESS2` (the current lack of health insurance among adults aged 18-64 years) for the state of Arizona. This function allows for multiple of these arguments.

```{r}

az_access <- get_places(state = "AZ", measure = "ACCESS2") 
head(az_access)


```

It is also worth noting that by default `geo` specifying geography is set to `county`. If instead we want to examine census tracts, we could specify the argument. Likewise, release is set to `2023`.

## Use Case

From here, we can start to have fun. It is fairly straight forward to begin exploring data. Here I will first filter out the data so that I can plot the age adjusted rates of lack of health insurance in Arizona.

Notice that the data provide you with confidence limits, so I have chosen to plot them here with error bars.

```{r}
az_access %>%
  filter(datavaluetypeid == "AgeAdjPrv") %>%
  ggplot(aes(data_value, reorder(locationname, data_value))) +
  geom_point(size = 2) +
  geom_errorbar(aes(xmin = low_confidence_limit, xmax = high_confidence_limit)) +
  labs(title = "Lack of health insurance among adults aged 18-64 years In Arizona Counties",
       y = "", x = "Percent") +
  theme_minimal() +
  theme(plot.title.position = "plot")
```

You can also extend this to multiple states to compare. You can easily query two (or more) state names, and plot them. Arizona seems to have a couple of counties that have a much higher rate compared to others.

```{r}

# multi state comparison
two <- get_places(state = c("AZ", "NV"), measure = "ACCESS2")

two %>%
  filter(datavaluetypeid == "AgeAdjPrv") %>%
  ggplot(aes(data_value, reorder(locationname, data_value), color = stateabbr)) +
  geom_point(size = 2) +
  geom_errorbar(aes(xmin = low_confidence_limit, xmax = high_confidence_limit)) +
  labs(title = 
         "Lack of health insurance among adults aged 18-64 years In Arizona and Nevada",
       y = "Counties", x = "Percent") +
  theme_minimal() +
  theme(plot.title.position = "plot")

```

We can go even further by comparing more states in the region. Here I have taken the average rate by state to easily compare. Texas appears to be far above the average.

```{r}
multi <- get_places(state = c("AZ", "NV", "NM", "TX", "CA"), measure = "ACCESS2") %>%
  filter(datavaluetypeid == "AgeAdjPrv") %>%
  summarise(.by = "stateabbr", mean_val = mean(data_value), mean_low = mean(low_confidence_limit), mean_high = mean(high_confidence_limit))

multi %>%
  ggplot(aes(mean_val, reorder(stateabbr, mean_val), color = stateabbr)) +
  geom_point(size = 2) +
  geom_errorbar(aes(xmin = mean_low, xmax = mean_high)) +
  labs(title = "Mean lack of health insurance among adults aged 18-64 years In Southwest States",
       y = "", x = "Percent") +
  theme_minimal() +
  theme(plot.title.position = "plot")

```


